---
title: "Snippet #3: Functions for simulating data"
author: "Giorgio Luciano"
date: "2023-03-4"
categories: 
  - R
  - tutorials
bibliography: references.bib
---

1.  Example of creating variables using `runif` and `rnorm`
2.  Writing a function that wraps all

First of all we use the `runif` and `rnorm` to have a look how they work.

```{r}

library(data.table)
x_min   <- 0
x_max   <- 10   
x_step  <- 0.01

y_mean  <- 0.5
y_sd    <- 0.25
y_min   <- -1
y_max   <- 1   

x       <- seq(x_min,x_max,x_step)
var_random  <- runif(x,y_min,y_max)
var_norm    <- rnorm(x,y_mean,y_sd) 

df  <- data.frame (x,var_random,var_norm)
dt  <- data.table(df)


```

```{r}

simpleDataset <- function(number_of_rows,means,sds)
{
l <- length(means)
res <- lapply(seq(1:l),function(x) 
	   eval(
	   parse(
	   text=paste("rnorm(",number_of_rows,",",means[x],",",sds[x],")",sep=""))
	   )
	   ) 
dat <- data.frame((sapply(res,c)))
id <- rownames(dat)
dat <-  cbind(id=id,dat)
dt <- data.table(dat)
return(dt)
}

```

Example 1: We simulate the values of the LDL cholesterol of *2* patients in *3* different times. The first one patient (X1) has an average value of *200* of LDL with a standard variation of *2* while the second (X2) has an average of *150* with a standard deviation of *10. Note: All values are expressed in mg/dL*

```{r}
dataset1 <- simpleDataset(3,c(200,180),c(2,10))
dataset1
```

Example 2: this time we combine `runif` and `simpleDataset`. We simulate the values of the LDL cholesterol of *5* patients in *7* different times. The values for each patient are between a *min = 100* and a *max = 150* with a standard deviation between a min sd = 10 and max sd = 40. We also simulate two time that presents outliers values between a *min = 180* and *max = 200* and an *min sd = 10* and *max sd = 40 .* We merge the values for each patient (*7* times + *2* outliers times) and finally we use the function melt to reshape the dataset.

```{r}
dat1 <- simpleDataset(number_of_rows=7,
					  means=runif(5,100,150),
				      sds=runif(5,10,40))
outliers <- simpleDataset(number_of_rows=2,
					  means=runif(5,180,200),
				      sds=runif(5,10,40))				  

dat1
outliers

dato     <-rbind(dat1,outliers) 
dt.melt <- melt(dat1, id.vars="id")
colnames(dt.melt) <- c("id","category","var1")
dt.melt$ncat <- as.numeric(dt.melt$category)

dt.melt
str(dt.melt)
```
