---
title: "Tip 4: boxplots and scatterplots: simple recipes"
author: "Giorgio Luciano"
date: "2023-03-11"
categories: 
  - R
  - tutorials
  - ggplot
  - visualization
  - recipes
bibliography: references.bib
---

**Summary**

1.  Simulate data, check and assign *data types*
2.  Create a scatterplot with `ggplot`
3.  Create violin plot with `ggstatsplot`

Example 1 we want to visualize the difference between two groups of patients that follow two different diets. *Group A* has an average of total cholesterol of *180* with a standard deviation of *20* while *Group B* and average of *200* with a standard deviation of 40

```{r}
library(MASS)
library(ggplot2)
library(data.table)


npatientsA <- 500
npatientsB <- 520
cholA <- mvrnorm(n=npatientsA, mu=180, Sigma=20, empirical=T)
cholB <- mvrnorm(n=npatientsB, mu=200, Sigma=40, empirical=T)

dataA <- cbind(cholA,rep("A",npatientsA))  
dataB <- cbind(cholB,rep("B",npatientsB))  

data <- data.frame(rbind(dataA,dataB))
colnames(data) <- c("Cholesterol","group")
data$Cholesterol <- as.numeric(data$Cholesterol)

p1 <-ggplot(data, aes(x = group, y = Cholesterol)) + geom_jitter(alpha=0.05) 

```

a few observation on the code. First of all, we need to input the data in a `data.frame` otherwise ggplot will give us an error. the second observation is that since we put `chr` labels on our groups we needed to define Cholesterol `as.numeric` in order to avoid strange results. Try to comment the line `data$Cholesterol <- as.numeric(data$Cholesterol)` and you can see by yourself what will happen. (hint: a "labelstorm!")

This very simple plot is one of my favorite. It let you see the raw data and immediately understand the distribution of your data and also avoid the pitfall of boxplot (see [@matejka2017])

If you need inferential statistics on your data another resource is [@ggstatsplot]. See the following example with our data. **NOTE** that we nee to transform the group label *as.factor*

```{r}
library(ggstatsplot)

data$group <- as.factor(data$group)

pstack  <- ggbetweenstats(data,group,Cholesterol)
                          
pstack    
```
